Choosing climate forcing for numerical simulations of the Cordilleran ice sheet

### 2013-02-21 sections/climate.tex
Added climate section foreword
Removed figure todonotes (update coming soon)
Fixed typos, language and style

\qiong[inline]{Include foreword. What is the forcing for?}

@@

WorldClim \citep{data:worldclim} is a high resolution climate dataset built from observation data. These climate surfaces were generated by spatial interpolation between a large set of weather stations using a 30\,arc\,s-aggregated flavour of hole-filled SRTM data over global land areas. This provides a resolution much higher than attained by general circulation models.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-temp}
	\end{center}
	\qiong[inline]{Use same order in text and figures}
	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
	\label{fig:temp}
\end{figure*}

Figure~\ref{fig:temp} shows the spatial distribution of winter (DJF) and summer (JJA) air surface temperatures in WorldClim. Temperatures generally decrease from south to north and regions further inland experience colder winters. It should be noted that the temperature gradient is much stronger in the winter than it is in the summer. At low elevations temperature get well above zero during the summer months over the entire modelling domain. In other words, there is a strong seasonality contrast beween coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-prec}
	\end{center}
	\qiong[inline]{Use same order in text and figures}
	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
	\label{fig:prec}
\end{figure*}

Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) and summer (JJA) precipitation rates in WorldClim. Coastal regions generally recieve much more precipitation than inland ones as a results of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascadia. In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Although coastal regions experience most precipitation during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative mean annual temperatures at present. In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model. Previous runs forced by annual means of temperature and precipitation not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.

The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. If good coverage exists in the southern parts of our modelling domain, several hundred kilometers can separate nearby stations in the north \citep{data:worldclim}. A second drawback of the dataset in an ice sheet modelling view is the lack of data on marine surfaces, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.

@@

In addition to the WorldClim data, we use surface air temperature and precipitation rate data from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR). Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} and monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean timeseries. Further information on the data used is gathered in Table \ref{tab:reanalyses}.

The spatial distribution of summer (JJA) air surface temperatures and winter (DJF) precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and are winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.

@@

Air surface temperature and precipitation rates from the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses and WorldClim data were reprojected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km. In addition, WorldClim data was extrapolated to marine surfaces using a nearest-neighbour approach. These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.

For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artifacts seen in figure~\ref{fig:prec}. This was done by averaging data locally in a circular neighbourhood of 7 pixels in diameter prior to reprojection. Yet we will show that this smoothing has very limited effect on the ice sheet model outcome. All the preprocessing steps described above were made in GRASS~GIS using scripts available on the first author's website.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-topo}
	\end{center}
	\qiong[inline]{Use same order in text and figures}
	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
	\label{fig:topo}
\end{figure*}

Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand. For reanalysis data, the circulation model surface topography (or surface geopotential height) was used as reference topography. For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography. As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset. The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}. A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations and no lapse-rate corrections apply to precipitation rates.

>>>>

We force an ice sheet model with temperature and precipitation data from one observational dataset, three global and one regional reanalyses. Here we present this climate data and describe how we use it as a forcing.

@@

WorldClim \citep{data:worldclim} is a high-resolution climate dataset built from observation data. These climate surfaces were generated by spatial interpolation between a large set of weather stations using a 30\,arc\,s-aggregated flavour of hole-filled SRTM data over global land areas. This provides a resolution much higher than attained by general circulation models.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-temp}
	\end{center}
	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
	\label{fig:temp}
\end{figure*}

Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim. Temperatures generally decrease from south to north and regions further inland experience colder winters. It should be noted that the temperature gradient is much stronger in the winter than it is in the summer. At low elevations, temperature get well above zero during the summer months over the entire modelling domain. In other words, there is a strong contrast in seasonality between coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-prec}
	\end{center}
	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
	\label{fig:prec}
\end{figure*}

Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) and summer (JJA) precipitation rates in WorldClim. Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Although coastal regions experience most precipitation during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative present mean annual temperatures. In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model. Previous runs forced by annual means of temperature and precipitation, not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.

The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate nearby stations in the north \citep{data:worldclim}. A second drawback of the dataset in an ice sheet modelling view is the lack of data on marine surfaces, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.

@@

Aside WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR). Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series. Further information on the data used is gathered in Table \ref{tab:reanalyses}.

The spatial distribution of summer (JJA) air surface temperatures and winter (DJF) precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.

@@

Air surface temperature and precipitation rates from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km. In addition, WorldClim data was extrapolated to marine surfaces using a nearest-neighbour approach. These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.

For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}. This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection. Yet we will show that this smoothing has very limited effect on the ice sheet model outcome. This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-topo}
	\end{center}
	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
	\label{fig:topo}
\end{figure*}

Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand. For reanalysis data, the circulation model surface topography (or surface geopotential height) was used as reference topography. For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography. As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset. The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}. A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections apply to precipitation rates.

### 2013-02-21 sections/model.tex
Added section 4 results (first shot)
Fixed typos, language and style

% ----------------------------------------------------------------------
\section{Model setup}
\label{sec:setup}
% ----------------------------------------------------------------------

We use PISM, a parallel ice sheet model (version stable 0.5.11). Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.

@@

Monthly temperatures and precipitation rates are converted to surface accumulation and ablation using a temperature-index (positive degree-day) model. Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\ask{Is this true? What is the current threshold in stable 0.5?} Ablation is computed from the number of positive degree-days, defined as the integrand of temperatures above 0\,\degree\,C in one year. Positive degree-days are computed by the \citet{calov-greve-2011} formula to account for day-to-day variability unrepresented in the monthly climate forcing. The ablation model use different degree-day factors for snow and ice and accounts for refreezing. Complete description of the ablation model is available in PISM documentation \citep[{\url{http://www.pism-docs.org}},][]{web:pism} and model parameters are explicited in table~\ref{tab:pdd}.

@@

	The number of positive degree-days, defined as the integrand of positive temperatures over time, is computed annually using a formula derived by \citet{calov-greve-2011} that accounts for day-to-day variability of temperature by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:

@@

A positive degree-day model is a very crude approximation of physical processes occurring at a glacier's surface. Yet we believe that it captures regional patterns of accumulation and ablation resonable enough for our study if accurate climate data is used, as we will show further.\ask{Is this paragraph really needed?}

@@

SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points. SIA velocities follow temperature-dependant creep. Temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions on 51 irregularly spaced layers in ice and 11 regularly spaced layers in rock. An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as lower boundary condition to the thermal model, and surface air temperature from the various climate forcing is used as upper boundary condition.

SSA velocities are determined by a a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescibed till friction angle$\phi$. We apply lower friction angles at low elevation to mimick the more slippery beds associated to the presence of marine sediments:

@@

In this study we model glacial inception and growth of the Cordilleran ice sheet to a state approaching its last glacial maximum. We mimick glacial conditions by applying constant temperature offsets homogeneously over the entire domain. Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval which we think to be representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref. Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.

>>>>

% ----------------------------------------------------------------------
\section{Model setup}
\label{sec:model}
% ----------------------------------------------------------------------

We use PISM, a parallel ice sheet model (version stable 0.5.11). Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012} and more extensively on-line \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.

@@

Monthly temperature and precipitation rate are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model. Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\ask{Is this true? What is the current threshold in stable 0.5?} Ablation is computed from the number of positive degree-days, defined as the integrand of temperatures above 0\,\degree\,C in one year. Positive degree-days are computed by the \citet{calov-greve-2011} formula to account for day-to-day variability, which is not represented in the monthly climate forcing. The ablation model use different degree-day factors for snow and ice and accounts for refreezing. Complete description of the ablation model is available in PISM documentation \citep[{\url{http://www.pism-docs.org}},][]{web:pism} and model parameters are expressed in table~\ref{tab:pdd}.

@@

	The number of positive degree-days is computed annually using a formula derived by \citet{calov-greve-2011} that accounts for day-to-day variability of temperature by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:

@@

A positive degree-day model is a very crude approximation of physical processes occurring at a glacier's surface. Yet we believe that it captures regional patterns of accumulation and ablation that are reasonable enough for our study if accurate climate data is used, as we will show further.\ask{Is this paragraph really needed?}

@@

SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points. SIA velocities follow temperature-dependant creep. Temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions on 51 irregularly spaced layers in ice and 11 regularly spaced layers in rock. An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as lower boundary condition to the thermal model, and the surface air temperature from the various climate forcing is used as upper boundary condition.

SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle$\phi$. We apply lower friction angles at low elevation to mimic the more slippery beds associated to the presence of marine sediments:

@@

In this study we model glacial inception and growth of the Cordilleran ice sheet to a state approaching its last glacial maximum. We mimic glacial conditions by applying constant temperature offsets homogeneously over the entire domain. Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref. Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.

### 2013-02-21 sections/results.tex
Starred figures and tables for two-columns output
Added more todonotes in figures
Removed figure todonotes (update coming soon)
Fixed typos, language and style

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-cool06}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\,\unit{\degree C} colder than present for each climate forcing.}
	\label{fig:cool06}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-best}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
	\label{fig:best}
\end{figure}

>>>>

Using climate forcing from WordClim data and the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-cool06}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
	\label{fig:cool06}
\end{figure*}

Figure~\ref{fig:cool06} shows the outcome of simulations run with a 6\degC temperature offset for each of the 6 climate forcings. In all simulations, ice accumulates and glaciers form as a result of the artificially cooled climate. Yet the magnitude of glaciation appears very different from one climate forcing to another. Whereas the NCAR and CFSR forcings produce a large ice sheet that covers most of the model domain land, the WorldClim, ERA-Interim and NARR forcings lead to more restrictive ice cover, bounded to mountain ice caps in the case of WorldClim data. The smoothing of CFSR precipitation data has limited visual effect on the resulting ice sheet geometry.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure*}

As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we ran simulations using offset values ranging from 2 to 9\degC for each climate forcing.

Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these~(48) simulations, grouped by climate forcing. It is again visible that different forcings generally lead to very different final ice cover. Notably, for the smallest temperature offset value of 2\degC that was used, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produced local ice caps only, restricted to the Wrangell-St.~Elias mountains in the case of WorldClim, an area presently glaciated. Several of the NCAR and CFSR simulations produced oversized ice-sheets whose extent is primirily bounded by the model domain boundary conditions rather than physical processes. Differences in regional patterns can also be noted. For instance ERA-Interim ice sheets generally appear more northernly-centered than NARR ice sheets.

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\todo[inline]{Use black-and-white-proof markers and highlight the ``best'' runs.}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}

Final ice volume and final ice coverage at the end of all simulations is quantified in Figure~\ref{fig:ivolarea}. The NCEP/NCAR and CFSR forcing lead to a much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings. The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-best}
	\end{center}
	\todo[inline]{The must would be to overlay geomorphic ice extent.}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
	\label{fig:best}
\end{figure*}

For each climate forcing, we selected the simulation that lead to a glaciated area closest to geomorphological reconstruction. This was done using the LGM extent countour from \needref shown in Figure~\ref{fig:locmap}, which covers an area of XXX \todo{Compute geomorpho-based ice cover area}. These ``best'' runs are presented in Figure~\ref{fig:best} along with the associated temperature offset values. We observe that although similar in size, the selected ice sheets are different in shape and flow patterns. More particularly, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles.\needref
